19Q1

### 2019-03-06 19Q1.Rmd
add 19Q1 cn and maf processed Rds

@@

This comprehensive cancer cell line information was curated by both Daniel Charytonowicz and myself.
```{r}
ccl_converter <- read.delim("./data_munging/cell_line_database_v1_20180911.tsv", row.names = 1, sep = "\t", header = TRUE)
lineage_converter <- unique(ccl_converter[, c("grouped_general_doid", "group_general_lineage_name")])
colnames(ccl_converter)[colnames(ccl_converter) == "Broad_ID"] <- "DepMap_ID"
```

@@

```{r}
ccl_info <- read.delim("./data_munging/19Q1/DepMap-2019Q1-celllines_20181115.csv", sep = ",", header = TRUE, na.strings = c("", NA))
ccl_info$Primary.Disease <- gsub("\\\\", "", ccl_info$Primary.Disease)
ccl_info$Primary.Disease <- gsub("Ewings", "Ewing's", ccl_info$Primary.Disease)
# From figshare
crispr_meta <- read.delim("./data_munging/19Q1/sample_info_19Q1_crispr_20181115.csv", sep = ",", header = TRUE, na.strings = c("", NA))
colnames(crispr_meta)[7] <- "CCLE_Name"
```

@@

Mutation calls (coding region, germline filtered):
```{r, eval=FALSE}
maf_raw <- read.delim("./data_munging/19Q1/depmap_19Q1_mutation_calls_20181114.csv.gz", header = TRUE, sep = ",", row.names = 1)
colnames(maf_raw)[colnames(maf_raw) == "Tumor_Sample_Barcode"] <- "CCLE_Name"
# Select columns
maf_df <- subset(maf_raw, select = c("Hugo_Symbol", "CCLE_Name", "DepMap_ID", "Variant_Classification", "Reference_Allele", "Tumor_Seq_Allele1", "Genome_Change"))
maf_df$Reference_Allele <- ifelse(maf_df$Reference_Allele == "-", 0, nchar(as.character(maf_df$Reference_Allele)))
maf_df$Tumor_Seq_Allele1 <- nchar(as.character(maf_df$Tumor_Seq_Allele1))
maf_df <- maf_df[maf_df$Variant_Classification != "Silent",]
# Add Mutation_Status column
maf_df$Mutation_Status_Nonsilent <- "Mutant"

@@

DepMap WES CN Data:
```{r, eval=FALSE}
cn <- read.delim("./data_munging/19Q1/public_19Q1_gene_cn_20181115.csv.gz", sep = ",", check.names = FALSE, header = TRUE)
# Remove Entrez gene IDs from colnames
# Format: Hugo_Symbol (Entrez ID)
colnames(cn) <- gsub(" .*", "", colnames(cn))
# Melt
cn_melt <- melt(data = cn, id.vars = "DepMap_ID", measure.vars = colnames(cn[2:ncol(cn)]), variable.name = "Hugo_Symbol", value.name = "CN")
saveRDS(cn_melt, "./data_munging/rds/cn_melt_19Q1.rds", compress = "xz")
```

@@

18Q3, Gene expression data (Reads Per Kilobase of transcript, per Million mapped reads, RPKM).
```{r, eval=FALSE}
ge_18Q3 <- read.delim("./../htscreens_giant_files/CCLE_DepMap_18q3_RNAseq_RPKM_20180718.gct.gz", skip = 2, header = TRUE, sep = "\t", check.names = FALSE)
# Edit columns
ge_18Q3$Name <- NULL
colnames(ge_18Q3)[1] <- "Hugo_Symbol"
# Melt
ge_melt_18Q3 <- melt(data = ge_18Q3, id.vars = "Hugo_Symbol", measure.vars = colnames(ge_18Q3[2:ncol(ge_18Q3)]), value.name = "RPKM")
## Split variable column
ge_melt_18Q3 <- with(ge_melt_18Q3, cbind(Hugo_Symbol, colsplit(variable, pattern = " ", names = c("CCLE_Name", "DepMap_ID")), RPKM))
## Remove parentheses around Broad IDs
ge_melt_18Q3$DepMap_ID <- gsub("\\(|\\)", "", ge_melt_18Q3$DepMap_ID)
saveRDS(ge_melt_18Q3, "./../htscreens_giant_files/ge_melt_18Q3.rds", compress = "xz")
```

19Q1, CCLE RNAseq gene expression data (log2(TPM+1)):
```{r, eval=FALSE}
ge_19Q1 <- read.delim("./../htscreens_giant_files/CCLE_depMap_19Q1_TPM_v2_20181115.csv.gz", header = TRUE, sep = ",", check.names = FALSE)
# Remove Entrez gene IDs from colnames
# Format: Hugo_Symbol (Entrez ID)
colnames(ge_19Q1) <- gsub(" .*", "", colnames(ge_19Q1))
colnames(ge_19Q1)[1] <- "DepMap_ID"
# Melt
ge_melt_19Q1_all <- melt(ge_19Q1, id.vars = "DepMap_ID", measure.vars = colnames(ge_19Q1)[2:ncol(ge_19Q1)], variable.name = "Hugo_Symbol", value.name = "TPM")
ge_melt_19Q1 <- unique(ge_melt_19Q1)
saveRDS(ge_melt_19Q1, "./data_munging/rds/ge_melt_19Q1.rds", compress = "xz")
```

>>>>

Note that the new 19Q1 release made multiple changes to variable names:

- They reverted back to "Broad_ID" from 18Q3 rather than "DepMap_ID" that was used in 18Q4
- Primary.Disease column of the cell line metadata file is more consistently named (Subtype.Disease now has the specific disease annotations), though there are a couple capitalization inconsistencies
- MAF file no longer has "CCLE_Name" column and has a duplicate column of the Broad IDs (DepMap_ID and Tumor_Sample_Barcode)

And data (the following is taken from the [data portal](https://depmap.org/portal/download/)):

- **Expression**: We use the GTEx pipeline. Since last release, there have been some updates to the versions of different components of the pipeline and minor bug fixes on GTEx's end.
- **Mutation**: In this release, we additionally process DepMap WES samples through the CGA mutation calling pipeline adapted for cell lines and available on FireCloud. We have added a column for the allele counts from this pipeline run (CGA_WES_AC). Previously, the WES_AC column was generated from running the Van Allen mutation calling pipeline for SNPs + Mutect2 for INDELs. This column is now named VA_WES_AC.
- **Copy Number**: We have moved to using the newest version of the GATK pipeline.

@@

This comprehensive cancer cell line information was curated by both Daniel Charytonowicz and myself.
```{r}
ccl_converter <- read.delim("./data_munging/cell_line_database_v1_20180911.tsv", row.names = 1, sep = "\t", header = TRUE)
lineage_converter <- unique(ccl_converter[, c("grouped_general_doid", "group_general_lineage_name")])
```

@@

19Q1 update fixed almost all of the inconsistencies of the Primary.Disease column, though there were come capitalization inconsistencies.
```{r}
ccl_info <- read.delim("./data_munging/19Q1/DepMap-2019q1-celllines_v2_20190305.csv", sep = ",", header = TRUE, na.strings = c("", NA))
# Convert Primary.Disease entries to all lower case to fix factorization
ccl_info$Primary.Disease <- as.factor(tolower(ccl_info$Primary.Disease))
# Harmonize ID column name
colnames(ccl_info)[colnames(ccl_info) == "DepMap_ID"] <- "Broad_ID"

# Sample info from figshare
crispr_meta <- read.delim("./data_munging/19Q1/sample_info_20190306.csv", sep = ",", header = TRUE, na.strings = c("", NA))
colnames(crispr_meta)[colnames(crispr_meta) == "CCLE_name"] <- "CCLE_Name"
```

@@

Mutation calls (coding region, germline filtered):
```{r, eval=FALSE}
maf_raw <- read.delim("./data_munging/19Q1/depmap_19Q1_mutation_calls_20190305.csv.gz", header = TRUE, sep = ",", row.names = 1)
# Harmonize ID column name
colnames(maf_raw)[colnames(maf_raw) == "DepMap_ID"] <- "Broad_ID"
# Select columns
maf_df <- subset(maf_raw, select = c("Hugo_Symbol", "Broad_ID", "Variant_Classification", "Reference_Allele", "Tumor_Seq_Allele1", "Genome_Change"))
maf_df$Reference_Allele <- ifelse(maf_df$Reference_Allele == "-", 0, nchar(as.character(maf_df$Reference_Allele)))
maf_df$Tumor_Seq_Allele1 <- nchar(as.character(maf_df$Tumor_Seq_Allele1))
maf_df <- maf_df[maf_df$Variant_Classification != "Silent",]
# Add Mutation_Status column
maf_df$Mutation_Status_Nonsilent <- "Mutant"

@@

DepMap WES CN Data:
```{r, eval=FALSE}
cn <- read.delim("./data_munging/19Q1/public_19Q1_gene_cn_20190305.csv.gz", sep = ",", check.names = FALSE, header = TRUE)
# Remove Entrez gene IDs from colnames, format is "Hugo_Symbol (Entrez ID)"
colnames(cn) <- gsub(" .*", "", colnames(cn))
colnames(cn)[1] <- "Broad_ID"
# Melt
cn_melt <- melt(data = cn, id.vars = "Broad_ID", measure.vars = colnames(cn[2:ncol(cn)]), variable.name = "Hugo_Symbol", value.name = "CN")
saveRDS(cn_melt, "./data_munging/rds/cn_melt_19Q1.rds", compress = "xz")
```

@@

18Q3, Gene expression data (Reads Per Kilobase of transcript, per Million mapped reads, RPKM).
```{r, eval=FALSE}
# ge_18Q3 <- read.delim("./../htscreens_giant_files/CCLE_depMap_19Q1_TPM_20190305.csv.gz", skip = 2, header = TRUE, sep = "\t", check.names = FALSE)
# # Edit columns
# ge_18Q3$Name <- NULL
# colnames(ge_18Q3)[1] <- "Hugo_Symbol"
# # Melt
# ge_melt_18Q3 <- melt(data = ge_18Q3, id.vars = "Hugo_Symbol", measure.vars = colnames(ge_18Q3[2:ncol(ge_18Q3)]), value.name = "RPKM")
# ## Split variable column
# ge_melt_18Q3 <- with(ge_melt_18Q3, cbind(Hugo_Symbol, colsplit(variable, pattern = " ", names = c("CCLE_Name", "DepMap_ID")), RPKM))
# ## Remove parentheses around Broad IDs
# ge_melt_18Q3$DepMap_ID <- gsub("\\(|\\)", "", ge_melt_18Q3$DepMap_ID)
# saveRDS(ge_melt_18Q3, "./../htscreens_giant_files/ge_melt_18Q3.rds", compress = "xz")
```

19Q1, CCLE RNAseq gene expression data (log2(TPM+1)):
```{r, eval=FALSE}
ge_19Q1 <- read.delim("./../htscreens_giant_files/CCLE_depMap_19Q1_TPM_20190305.csv.gz", header = TRUE, sep = ",", check.names = FALSE)
# Remove Entrez gene IDs from colnames, format is "Hugo_Symbol (Entrez ID)"
colnames(ge_19Q1) <- gsub(" .*", "", colnames(ge_19Q1))
colnames(ge_19Q1)[1] <- "Broad_ID"
# Melt
ge_melt_19Q1_all <- melt(ge_19Q1, id.vars = "Broad_ID", measure.vars = colnames(ge_19Q1)[2:ncol(ge_19Q1)], variable.name = "Hugo_Symbol", value.name = "TPM")
# ge_melt_19Q1 <- unique(ge_melt_19Q1_all)
saveRDS(ge_melt_19Q1, "./data_munging/rds/ge_melt_19Q1.rds", compress = "xz")
```
